---
title: "CITE-seq Pilot"
output: github_document
---

```{r setup, include=FALSE}
unix::rlimit_as(1e12)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, 
                      warning = FALSE, comment = "#", dpi = 300)
options(width = 999)
```

## Packages

```{r}
# single-cell data analysis
library(Seurat)
library(miQC)
library(scater)
library(MCPcounter)

# Gene Ontology
library(clusterProfiler)
library(org.Hs.eg.db)

# Data wrangling
library(tidyverse)

# Plotting
library(tidytext)
library(UpSetR)
library(ggridges)
library(RColorBrewer)
library(scico)
library(ggsci)
library(cowplot)
```

```{r, echo=FALSE}
features_df <-
    "/lab-share/IM-Gutierrez-e2/Public/Lab_datasets/CITEseq_pilot_2/SN0257788/broad/hptmp/curtism/bwh10x/KW10170_Maria/220617_10X_KW10170_bcl/cellranger-6.1.1/GRCh38/BRI-1743_hashing/outs/filtered_feature_bc_matrix/features.tsv.gz" %>%
    read_tsv(col_names = c("gene_id", "gene_name", "phenotype"))

demuxlet_sng <- "./demuxlet/demuxlet_allsnps.best" %>%
    read_tsv() %>%
    select(barcode = BARCODE, best = BEST) %>%
    extract(best, c("status", "sample"), "([^-]+)-(.+)") %>%
    filter(status == "SNG") %>%
    select(barcode, sample)

gene_ids <- features_df %>%
    filter(phenotype == "Gene Expression") %>%
    select(gene_id, gene_name)

stims <- c("day0", "IL4 24hr", "BCR 24hr", "BCR 72hr", "TLR7 24hr", "TLR7 72hr",
           "BCR+TLR7 24hr", "BCR+TLR7 72hr", "CpG 24hr", "CpG 72hr", 
           "sCD40L 24hr", "sCD40L 72hr", "DN2 24hr", "DN2 72hr")

stims_colors <- c("grey", "black", brewer.pal(12, "Paired")) %>%
    setNames(stims)
```

## QC and data cleaning

This process is describe in the README file in the parent directory.

Briefly, we used a custom demultiplexing approach based on HTO counts. We extracted cells classified by singlets by both this approach, Scrublet, and demuxlet.

```{r}
bcells_singlet <- readRDS("../data/pilot2.Rdata")
```


## PCA

```{r}
bcells_singlet <- bcells_singlet %>%
    FindVariableFeatures(nfeatures = 2000, selection.method = "vst") %>%
    ScaleData(., features = rownames(.)) %>%
    RunPCA(., features = VariableFeatures(.))
```


```{r, echo=FALSE, fig.height=7}
meta_df <- bcells_singlet@meta.data %>%
    as_tibble(rownames = "barcode") %>%
    select(barcode, stim = HTO_maxID, n_genes = nFeature_RNA)

pca_cell_emb <- bcells_singlet@reductions$pca@cell.embeddings %>%
    as_tibble(rownames = "barcode") %>%
    left_join(meta_df) %>%
    select(barcode, stim, n_genes, PC_1:PC_4)

pca_plot <- ggplot(pca_cell_emb, aes(PC_1, PC_2)) +
  geom_point(aes(color = stim), size = .5) +
  scale_color_manual(values = stims_colors) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.key.height = unit(.75, "lines")) +
  guides(color = guide_legend(override.aes = list(size = 2)))

sdev_plot <- tibble(sdev = bcells_singlet@reductions$pca@stdev) %>%
    rownames_to_column("pc") %>%
    mutate(pc = fct_inorder(pc)) %>%
    filter(pc %in% 1:25) %>%
    ggplot(aes(pc, sdev)) +
    geom_line(aes(group = 1)) +
    scale_x_discrete(breaks = c(1, seq(5, 25, 5))) +
    theme_bw() +
    theme(panel.grid.minor = element_blank()) +
    labs(x = "Principal component", y = "Standard deviation")

pca_df <- bcells_singlet@reductions$pca@feature.loadings %>%
    as_tibble(rownames = "gene_id")

loadings_plot <- pca_df %>%
    select(gene_id, PC_1, PC_2) %>%
    pivot_longer(-gene_id, names_to = "pc") %>%
    mutate(direction = case_when(sign(value) == 1 ~ "+",
				 sign(value) == -1 ~ "-",
				 sign(value) == 0 ~ "0")) %>%
    group_by(pc, direction) %>%
    top_n(15, abs(value)) %>%
    ungroup() %>%
    left_join(gene_ids) %>%
    ggplot(aes(value, reorder_within(gene_name, value, pc))) +
    geom_col(aes(fill = direction), show.legend = FALSE) +
    scale_y_discrete(labels = function(x) sub("^([^_]+).*$", "\\1", x)) +
    scale_fill_manual(values = c("+" = "cornflowerblue", "-" = "tomato3")) +
    facet_wrap(~pc, scales = "free") +
    labs(x = "Loading", y = NULL) +
    theme_bw()

plot_grid(plot_grid(pca_plot, sdev_plot, rel_widths = c(1, .7)),
          loadings_plot,
          ncol = 1, rel_heights = c(.7, 1))
```


### Number of genes and cells per individual

```{r, echo=FALSE, fig.width=8, fig.height=3}
pca_samples <- left_join(pca_cell_emb, demuxlet_sng) %>%
    ggplot(aes(PC_1, PC_2)) +
    geom_point(aes(color = sample), size = .5) +
    scale_color_npg() +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    labs(color = "Individual") +
    guides(color = guide_legend(override.aes = list(size = 2)))

pca_ngenes <- pca_cell_emb %>%
    ggplot(aes(PC_1, PC_2)) +
    geom_point(aes(fill = n_genes, color = n_genes), size = .5, shape = 19) +
    scale_color_viridis_c(guide = "none") +
    scale_fill_viridis_c(guide = guide_colorbar(barwidth = .5)) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    labs(fill = "Number of\ngenes")

plot_grid(pca_ngenes, pca_samples, nrow = 1)
```

```{r, echo=FALSE, fig.height=4, fig.width=10}
left_join(meta_df, demuxlet_sng) %>%
  count(stim, sample) %>%
  mutate(stim = factor(stim, levels = stims)) %>%
  ggplot(aes(stim, n, fill = sample)) +
  geom_col(color = "black", size = .5, position = "dodge") +
  scale_fill_npg() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.x = element_blank()) +
  labs(x = NULL, y = "Number of droplets")
```


## UMAP and clustering

```{r}
bcells_singlet <- bcells_singlet %>%
  RunUMAP(dims = 1:20, verbose = FALSE) %>%
  FindNeighbors(dims = 1:20, verbose = FALSE) %>%
  FindClusters(resolution = 0.5, verbose = FALSE)
```


```{r, echo = FALSE, fig.height=3, fig.width=10}
meta_data <- bcells_singlet@meta.data %>%
    as_tibble(rownames = "barcode") %>%
    select(barcode, cluster = seurat_clusters, stim = HTO_maxID, 
	   percent_mt, percent_ribo, n_genes = nFeature_RNA, umi = nCount_RNA) %>%
    mutate(stim = factor(stim, levels = stims))

umap_df <- as.data.frame(bcells_singlet@reductions$umap@cell.embeddings) %>%
    as_tibble(rownames = "barcode") %>%
    left_join(meta_data)

umap_stims <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = stim)) +
    geom_point(size = .5) +
    scale_color_manual(values = stims_colors) +
    theme_bw() +
    theme(panel.grid = element_blank(),
          panel.border = element_blank(),
          axis.line = element_blank(),
          axis.ticks = element_blank(),
          legend.key.height = unit(.75, "lines")) +
    guides(color = guide_legend(override.aes = list(size = 2)))

cluster_labels <- umap_df %>%
  group_by(cluster) %>%
  summarise_at(vars(UMAP_1, UMAP_2), median) %>%
  ungroup()

umap_clusters <- ggplot(umap_df) +
    geom_point(aes(UMAP_1, UMAP_2, color = cluster), size = .5) +
    geom_label(data = cluster_labels, 
               aes(UMAP_1, UMAP_2, label = cluster),
               alpha = .5) +
    scale_color_viridis_d() +
    theme_bw() +
    theme(panel.grid = element_blank(),
          panel.border = element_blank(),
          axis.line = element_blank(),
          axis.ticks = element_blank(),
          legend.key.height = unit(.75, "lines")) +
    guides(color = guide_legend(override.aes = list(size = 2)))

plot_grid(umap_stims, umap_clusters, nrow = 1)
```


```{r, echo = FALSE, fig.width=5, fig.height=3}
umap_df %>%
  left_join(demuxlet_sng) %>%
  select(barcode, sample, UMAP_1, UMAP_2) %>%
  ggplot() +
    geom_point(aes(UMAP_1, UMAP_2, color = sample), size = .5) +
    scale_color_npg() +
    theme_bw() +
    theme(panel.grid = element_blank(),
          panel.border = element_blank(),
          axis.line = element_blank(),
          axis.ticks = element_blank()) +
    guides(color = guide_legend(override.aes = list(size = 2)))
```


### Hashtag staining

```{r, echo=FALSE, fig.width=10, fig.height=10}
hash_rawcounts <- t(bcells_singlet@assays$HTO@counts) %>%
  as_tibble(rownames = "barcode") %>%
  pivot_longer(-barcode, names_to = "hash") %>%
  mutate(hash = factor(hash, levels = str_sort(unique(hash), numeric = TRUE)),
         value = log10(value),
         value = ifelse(value == -Inf, 0, value)) %>%
  as_tibble()

umap_df %>%
    left_join(hash_rawcounts, by = "barcode") %>%
    select(barcode, UMAP_1, UMAP_2, hash, value) %>%
    ggplot(aes(UMAP_1, UMAP_2, color = value, fill = value), shape = 19) +
    geom_point(size = .1) +
    scale_color_viridis_c(guide = "none") +
    scale_fill_viridis_c(guide = guide_colorbar(barwidth = .5)) +
    facet_wrap(~hash, ncol = 3) +
    theme_bw() +
    theme(panel.grid = element_blank(),
	  panel.border = element_blank(),
	  axis.line = element_blank(),
	  axis.ticks = element_blank(),
	  axis.title = element_blank(),
	  axis.text = element_blank()) +
    labs(fill = "log10\n(raw HTO counts)")
```

### Cell type annotation with MCP counter

```{r, echo = FALSE, fig.width=9, fig.height=6}
mcp_scores <- 
  MCPcounter.estimate(expression = bcells_singlet@assays$RNA@data,
                      featuresType = "ENSEMBL_ID") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("barcode") %>%
  as_tibble() %>%
  pivot_longer(-barcode, names_to = "cell_type")

umap_celltype_list <- umap_df %>%
  left_join(mcp_scores, by = "barcode") %>%
  split(.$cell_type) %>%
  map(~ggplot(.) +
  geom_point(aes(UMAP_1, UMAP_2, color = value, fill = value), size = .1, shape = 19) +
  scale_color_viridis_c(guide = "none") +
  scale_fill_viridis_c("", guide = guide_colorbar(barwidth = .5)) +
  facet_wrap(~cell_type, scales = "free", ncol = 3) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        legend.key.height = unit(.75, "lines"),
        axis.title = element_blank(),
        axis.text = element_blank()))

plot_grid(plotlist = umap_celltype_list, ncol = 3)
```

### Total of B and non-B cells

```{r, echo=FALSE, fig.width=8, fig.height=4}
umap1_side <- umap_df %>%
    select(barcode, UMAP_1) %>%
    left_join(mcp_scores, by = "barcode") %>%
    filter(cell_type == "B lineage") %>%
    group_by(i = UMAP_1 >= 0) %>%
    summarise(m = mean(value)) %>%
    ungroup() %>%
    filter(m == max(m)) %>%
    pull(i)

cell_type_class <- umap_df %>%
    mutate(cell_type_cluster = case_when(umap1_side == FALSE & UMAP_1 < 0 ~ "B cell",
					 umap1_side == FALSE & UMAP_1 >= 0 ~ "non-B cell",
					 umap1_side == TRUE & UMAP_1 >= 0 ~ "B cell",
					 umap1_side == TRUE & UMAP_1 < 0 ~ "non-B cell"))

cell_totals <- cell_type_class %>%
    count(cell_type_cluster, stim)

cells_p1 <- cell_totals %>%
    group_by(cell_type_cluster) %>%
    summarise(n = sum(n)) %>%
    ungroup() %>%
    mutate(cell_type_cluster = str_pad(cell_type_cluster, width = 13, side = "left"),
	   time = "Total") %>%
    ggplot(aes(cell_type_cluster, n, fill = cell_type_cluster)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    scale_fill_npg() +
    facet_wrap(~time) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = NULL, y = "Number of cells")

cells_p2 <- cell_totals %>%
    separate(stim, c("stim", "time"), sep = " ") %>%
    mutate(time = ifelse(is.na(time) & stim == "day0", "0hr", time),
	 stim = fct_inorder(stim),
	 time = factor(time, levels = c("0hr", "24hr", "72hr"))) %>%
    ggplot(aes(stim, n, fill = cell_type_cluster)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_npg() +
    facet_grid(.~time, scales = "free_x", space = "free") +
    theme_bw() +
    theme(panel.grid.major.x = element_blank(),
	axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = NULL, y = "Number of cells", fill = NULL)

plot_grid(cells_p1, cells_p2, rel_widths = c(.25, 1))
```




### Cell cycling

```{r, echo=FALSE, fig.height=5, fig.width=8}
mki_gene_df <- gene_ids %>%
    filter(gene_name == "MKI67") %>%
    select(gene_id, gene_name)

mki_gene_quant <- bcells_singlet@assays$RNA@data %>% 
    .[mki_gene_df$gene_id, ,drop = FALSE] %>%
    as_tibble(rownames = "gene_id") %>%
    left_join(mki_gene_df, by = "gene_id") %>%
    pivot_longer(-c("gene_id", "gene_name"), 
                 names_to = "barcode", values_to = "gene_exp")

ki67 <- umap_df %>%
    select(barcode, UMAP_1, UMAP_2) %>%
    left_join(mki_gene_quant, by = "barcode") %>%
    ggplot(aes(UMAP_1, UMAP_2, color = gene_exp, fill = gene_exp)) +
    geom_point(size = .25, shape = 19) +
    scale_color_viridis_c(guide = "none") +
    scale_fill_viridis_c(guide = guide_colorbar(barwidth = .5)) +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          axis.title = element_blank(),
          axis.text = element_blank()) +
    labs(title = "MKI67 RNA expression", fill = NULL)

ngenes <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = n_genes, fill = n_genes)) +
    geom_point(size = .25, shape = 19) +
    scale_color_viridis_c(guide = "none") +
    scale_fill_viridis_c(guide = guide_colorbar(barwidth = .5)) +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          axis.title = element_blank(),
          axis.text = element_blank()) +
    labs(title = "Number of genes", fill = NULL)

mt <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = percent_mt, fill = percent_mt)) +
    geom_point(size = .25, shape = 19) +
    scale_color_viridis_c(guide = "none") +
    scale_fill_viridis_c(guide = guide_colorbar(barwidth = .5)) +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          axis.title = element_blank(),
          axis.text = element_blank()) +
    labs(title = "% UMIs from mitochondrial genes", fill = NULL)

ribo <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = percent_ribo, fill = percent_ribo)) +
    geom_point(size = .25, shape = 19) +
    scale_color_viridis_c(guide = "none") +
    scale_fill_viridis_c(guide = guide_colorbar(barwidth = .5)) +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          axis.title = element_blank(),
          axis.text = element_blank()) +
    labs(title = "% UMIs from ribosomal genes", fill = NULL)


plot_grid(ki67, ngenes, mt, ribo, nrow = 2)
```


## Extract B cells

```{r}
stims_24 <- c("day0", "IL4 24hr", "BCR 24hr", "TLR7 24hr", "BCR+TLR7 24hr", "DN2 24hr")

filtered_bcells <- cell_type_class %>%
  filter(stim %in% stims_24, cell_type_cluster == "B cell")

bcells_filt <- bcells_singlet %>%
  subset(cells = filtered_bcells$barcode)
```


```{r}
bcells_filt <- bcells_filt %>%
    FindVariableFeatures(nfeatures = 1000, selection.method = "vst") %>%
    ScaleData(., features = rownames(.)) %>%
    RunPCA(., features = VariableFeatures(.)) %>%
    RunUMAP(dims = 1:20, verbose = FALSE) %>%
    FindNeighbors(dims = 1:20, verbose = FALSE) %>%
    FindClusters(resolution = 0.5, verbose = FALSE)
```


```{r, echo=FALSE, fig.height=3, fig.width=10}
pca_emb_filt <- bcells_filt@reductions$pca@cell.embeddings %>%
    as_tibble(rownames = "barcode") %>%
    left_join(meta_df) %>%
    select(barcode, stim, PC_1:PC_4)

umap_filt_df <- as.data.frame(bcells_filt@reductions$umap@cell.embeddings) %>%
    as_tibble(rownames = "barcode") %>%
    left_join(meta_data)

pca_filt_plot <- ggplot(pca_emb_filt, aes(PC_1, PC_2)) +
  geom_point(aes(fill = stim), size = 1.5, shape = 21, color = "grey80") +
  scale_fill_manual(values = stims_colors[stims_24]) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.key.height = unit(.75, "lines")) +
  guides(fill = guide_legend(override.aes = list(size = 2)))

umap_filt_plot <- ggplot(umap_filt_df, aes(UMAP_1, UMAP_2, fill = stim)) +
    geom_point(size = 1.5, shape = 21, color = "grey80") +
    scale_fill_manual(values = stims_colors[stims_24]) +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          axis.line = element_blank(),
          axis.ticks = element_blank(),
          legend.key.height = unit(.75, "lines")) +
    guides(fill = guide_legend(override.aes = list(size = 2)))

plot_grid(pca_filt_plot + theme(legend.position = "none"), 
          umap_filt_plot + theme(legend.position = "none"),
          get_legend(umap_filt_plot), 
          nrow = 1,
          rel_widths = c(1, 1, .25))
```

### Marker genes at 24 hours in respect to day 0

```{r, echo = FALSE}
plot_markers <- function(cluster_df, seurat_obj) {
  
  top_markers <- cluster_df %>%
    as_tibble() %>%
    left_join(select(features_df, 1:2), by = c("gene" = "gene_id")) %>%
    group_by(cluster) %>%
    top_n(15, avg_log2FC) %>%
    ungroup() %>%
    select(cluster, gene_name, gene_id = gene, avg_log2FC)
  
  cluster_cell <- 
    tibble(cluster_cell = Idents(seurat_obj),
           barcode = names(Idents(seurat_obj))) %>%
    mutate(cluster_cell = factor(cluster_cell, levels = levels(top_markers$cluster)))
  
  cell_gene_expr <- seurat_obj@assays$RNA@data %>%
    .[unique(top_markers$gene_id), ] %>%
    as.data.frame() %>%
    rownames_to_column("gene_id") %>%
    as_tibble() %>%
    pivot_longer(-gene_id, names_to = "barcode", values_to = "logexpr")
  
  top_marker_expr <- cell_gene_expr %>%
    left_join(cluster_cell, by = "barcode") %>%
    inner_join(top_markers, by = "gene_id")
  
  top_marker_summary <- top_marker_expr %>%
    group_by(cluster = cluster_cell, gene_name) %>%
    summarise(prop_expr = mean(logexpr > 0),
              scale_expr = mean(logexpr)) %>%
    ungroup() %>%
    left_join(select(top_markers, cluster_top = cluster, gene_name, avg_log2FC)) %>%
    mutate_at(vars(cluster, cluster_top), factor) %>%
    arrange(cluster_top, avg_log2FC) %>%
    mutate(gene_name = fct_inorder(gene_name)) 
  
  
  ggplot(top_marker_summary, aes(cluster, gene_name)) +
    geom_point(aes(size = prop_expr, fill = scale_expr), 
               color = "black", shape = 21) +
    scale_size(range = c(0.1, 4), labels = scales::percent) +
    scale_fill_viridis_c(option = "magma") +
    facet_wrap(~cluster_top, scales = "free", ncol = 3) +
    theme_bw() +
    theme(axis.line = element_blank(),
          axis.text = element_text(size = 8),
          panel.grid = element_line(color = "grey96"),
          panel.border = element_blank(),
          legend.position = "top") +
    labs(x = NULL, y = NULL, 
         fill = "Scaled\nExpression", 
         size = "% of cells") +
    guides(fill = guide_colorbar(barheight = .5))
}
```

### Marker genes across all conditions at 24h and day 0


```{r}
Idents(bcells_filt) <- "HTO_maxID"

all_markers_df <- 
  FindAllMarkers(bcells_filt,
                 only.pos = TRUE,
                 min.pct = 0.1,
                 logfc.threshold = 1) %>%
  as_tibble() %>%
  filter(p_val_adj < 0.05)
```

```{r, echo = FALSE, fig.width=10, fig.height=5}
all_markers_df %>%
  mutate(cluster = factor(cluster, levels = stims_24)) %>%
  plot_markers(bcells_filt) +
  scale_x_discrete(labels = function(x) sub("\\s", "\n", x))
```

```{r}
il4_markers <- 
    FindMarkers(bcells_filt, 
                   ident.1 = "IL4 24hr",
                   ident.2 = "day0",
                   only.pos = TRUE,
                   min.pct = 0.1,
                   logfc.threshold = 0.5) %>%
    rownames_to_column("gene") %>%
    as_tibble()

bcr_markers <- 
    FindMarkers(bcells_filt, 
                   ident.1 = "BCR 24hr",
                   ident.2 = "day0",
                   only.pos = TRUE,
                   min.pct = 0.1,
                   logfc.threshold = 0.5) %>%
    rownames_to_column("gene") %>%
    as_tibble()

tlr_markers <- 
    FindMarkers(bcells_filt, 
                   ident.1 = "TLR7 24hr",
                   ident.2 = "day0",
                   only.pos = TRUE,
                   min.pct = 0.1,
                   logfc.threshold = 0.5) %>%
    rownames_to_column("gene") %>%
    as_tibble()

bcrtlr_markers <- 
    FindMarkers(bcells_filt, 
                   ident.1 = "BCR+TLR7 24hr",
                   ident.2 = "day0",
                   only.pos = TRUE,
                   min.pct = 0.1,
                   logfc.threshold = 0.5) %>%
    rownames_to_column("gene") %>%
    as_tibble()


dn2_markers <- 
    FindMarkers(bcells_filt, 
                   ident.1 = "DN2 24hr",
                   ident.2 = "day0",
                   only.pos = TRUE,
                   min.pct = 0.1,
                   logfc.threshold = 0.5) %>%
    rownames_to_column("gene") %>%
    as_tibble()

markers_df <-
  bind_rows("IL4" = il4_markers,
	    "BCR" = bcr_markers,
            "TLR7" = tlr_markers,
            "BCR+TLR7" = bcrtlr_markers,
            "DN2" = dn2_markers,
            .id = "stim") %>%
  filter(p_val_adj < 0.05) %>%
  select(-p_val, -p_val_adj)
```


```{r, echo = FALSE, fig.height=5, fig.width=6.5}
top30_markers <- markers_df %>%
  group_by(stim) %>%
  top_n(30, avg_log2FC) %>%
  ungroup() %>%
  pivot_longer(pct.1:pct.2, names_to = "condition", values_to = "perc") %>%
  mutate(condition = ifelse(condition == "pct.1", stim, "day0"))

bcells_filt_subset <- bcells_filt %>%
  subset(idents = stims_24)

meta_data_filtsub <- bcells_filt_subset@meta.data %>% 
    as_tibble(rownames = "barcode") %>%
    select(barcode, stim = HTO_maxID)

filt_stim_avg_expr <- bcells_filt_subset@assays$RNA@data %>%
    .[unique(top30_markers$gene), ] %>%
    as_tibble(rownames = "gene") %>%
    pivot_longer(-gene, names_to = "barcode") %>%
    left_join(meta_data_filtsub) %>%
    mutate(stim = sub("\\s24hr", "", stim)) %>%
    filter(value > 0) %>%
    group_by(stim, gene) %>%
    summarise(value = mean(value)) %>%
    ungroup() 

markers_plot_df <- top30_markers %>%
  left_join(filt_stim_avg_expr, by = c("condition" = "stim", "gene")) %>%
  left_join(gene_ids, by = c("gene" = "gene_id")) %>%
  mutate(condition = fct_relevel(condition, "day0", after = Inf),
	 stim = fct_inorder(stim))


ggplot(markers_plot_df, aes(x = condition, y = reorder_within(gene_name, avg_log2FC, stim))) +
    geom_point(aes(size = perc, fill = value), 
               color = "black", shape = 21) +
    scale_size(range = c(0.1, 3.5), labels = scales::percent) +
    scale_fill_viridis_c(option = "magma") +
    scale_y_discrete(labels = function(x) sub("^([^_]+).*$", "\\1", x)) +
    facet_wrap(~stim, scales = "free", nrow = 1) +
    theme_bw() +
    theme(axis.line = element_blank(),
          axis.text = element_text(size = 7),
          panel.grid = element_line(color = "grey96"),
          panel.border = element_blank(),
          legend.position = "top") +
    labs(x = NULL, y = NULL, 
         fill = "Scaled\nExpression", 
         size = "% of cells") +
    guides(fill = guide_colorbar(barheight = .5))
  
```


```{r, echo=FALSE, fig.width=6.5, fig.height=5}
input_list <- markers_df %>%
  split(.$stim) %>%
  map("gene")

upset(fromList(input_list), order.by = "freq")
```

### Gene ontology analysis of shared and exclusive genes

```{r, echo = FALSE}
run_enrichment <- function(gene_list) {

  enrichGO(gene = gene_list,
           OrgDb = org.Hs.eg.db,
           ont = "BP",
           keyType = "ENSEMBL",
           pAdjustMethod = "fdr",
           pvalueCutoff = 0.05,
           qvalueCutoff = 0.05,
           readable = TRUE) %>%
    as.data.frame() %>%
    as_tibble()
}

enrich_df <- markers_df %>%
    split(.$stim) %>%
    map("gene") %>%
    map_df(run_enrichment, .id = "set")
```

```{r, echo = FALSE, fig.width=10, fig.height=12}
go_top <- enrich_df %>%
    group_by(set) %>%
    top_n(30, -log10(qvalue)) %>%
    ungroup() %>%
    mutate(set = factor(set, levels = c("IL4", "BCR", "TLR7", "BCR+TLR7", "DN2"))) %>%
    arrange(set, qvalue)

goterms <- go_top %>%
  arrange(qvalue) %>%
  distinct(Description) %>%
  pull(Description)

go_top %>%
    mutate(Description = factor(Description, levels = rev(goterms))) %>%
    ggplot(aes(set, Description)) +
        geom_point(aes(size = Count, fill = -log10(qvalue)), shape = 21) +
        scale_size(range = c(1, 5)) +
        scale_fill_viridis_c(option = "magma") +
        scale_y_discrete(labels = function(x) str_trunc(x, width = 80)) +
        theme_bw() +
        theme(axis.title = element_blank(),
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 11, hjust = 1),
              axis.ticks = element_blank(),
              panel.grid.major.x = element_blank(),
              panel.border = element_blank(),
              legend.position = "top",
              legend.justification = c(1, 0)) +
        labs(size = "# Genes:") +
        guides(fill = guide_colorbar(direction = "horizontal", barheight = 0.5),
               size = guide_legend(direction = "horizontal"))

```



## ADTs

```{r, echo=FALSE, fig.height=55, fig.width=9}
adt_quants <- bcells_singlet@assays$ADT@data %>%
    as_tibble(rownames = "ab") %>%
    pivot_longer(-ab, names_to = "barcode", values_to = "ab_level") %>%
    mutate(ab = factor(ab, levels = str_sort(unique(ab), numeric = TRUE)))

# shrink outliers
adt_quants_norm <- adt_quants %>%
  group_by(ab) %>%
  mutate(q01 = quantile(ab_level, 0.01),
         q99 = quantile(ab_level, 0.99),
         ab_level = case_when(ab_level < q01 ~ q01,
                              ab_level > q99 ~ q99,
                              TRUE ~ ab_level)) %>%
  ungroup()

bcell_prots_plot_list <- umap_df %>%
    select(barcode, UMAP_1, UMAP_2) %>%
    left_join(adt_quants_norm, by = "barcode") %>%
    split(.$ab) %>%
    map(~ggplot(data = ., aes(UMAP_1, UMAP_2, color = ab_level)) +
            geom_point(size = .2) +
	    scale_color_scico(palette = "lajolla",
			      labels = function(x) str_pad(x, 3),
			      guide = guide_colorbar(barwidth = .5,
						     barheight = 3)) +
            facet_wrap(~ab) +
            theme_bw() +
            theme(panel.grid = element_blank(),
                  panel.border = element_blank(),
                  axis.title = element_blank(),
                  axis.text = element_blank(),
                  strip.background = element_blank(),
                  axis.line = element_blank(),
                  axis.ticks = element_blank()) +
            labs(color = NULL))

plot_grid(plotlist = bcell_prots_plot_list, ncol = 3)
```

