---
title: "Read me"
output: github_document
---

```{r, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 300)
```


```{r}
library(tidyverse)
library(tidytext)
library(cowplot)
```

```{r}
quant_df <- read_tsv("./compiled_expression.tsv") %>%
    mutate(gene_id = str_remove(gene_id, "\\.\\d+$"))

expressed_df <- quant_df %>%
    group_by(gene_id) %>%
    filter(mean(tpm > 1) > .5) %>%
    ungroup()

#variable_genes <- expressed_df %>%
#    group_by(gene_id) %>%
#    summarise(v = var(tpm)) %>%
#    ungroup() %>%
#    arrange(desc(v)) %>%
#    slice(1:5000)

expressed_matrix <- expressed_df %>%
    #filter(gene_id %in% variable_genes$gene_id) %>%
    select(-gene_name) %>%
    pivot_wider(names_from = gene_id, values_from = tpm) %>%
    column_to_rownames("sampleid") %>%
    as.matrix()

gene_ids <- distinct(expressed_df, gene_id, gene_name)
```


```{r}
ifn_a <- 
    "https://www.gsea-msigdb.org/gsea/msigdb/download_geneset.jsp?geneSetName=HALLMARK_INTERFERON_ALPHA_RESPONSE&fileType=grp" %>%
    read_lines() %>%
    discard(grepl("^HALLMARK|^#", .))

ifn_g <- 
    "https://www.gsea-msigdb.org/gsea/msigdb/download_geneset.jsp?geneSetName=HALLMARK_INTERFERON_GAMMA_RESPONSE&fileType=grp" %>%
    read_lines() %>%
    discard(grepl("^HALLMARK|^#", .))

ifn_davenport <- c("HERC5", "IFI27", "IRF7", "ISG15", "LY6E", "MX1", 
		   "OAS2", "OAS3", "RSAD2", "USP18", "GBP5")

ifn_expression <- expressed_df %>%
    filter(gene_name %in% c(ifn_a, ifn_g, ifn_davenport)) %>%
    group_by(gene_id) %>%
    mutate(stdexp = GenABEL::rntransform(tpm),
	   stdscaled = stdexp + abs(min(stdexp))) %>%
    ungroup()

ifn_a_scores <- ifn_expression %>%
    filter(gene_name %in% ifn_a) %>%
    group_by(sampleid) %>%
    summarise(ifn_a = sum(stdscaled)) %>%
    ungroup()

ifn_g_scores <- ifn_expression %>%
    filter(gene_name %in% ifn_g) %>%
    group_by(sampleid) %>%
    summarise(ifn_g = sum(stdscaled)) %>%
    ungroup()

ifn_davenp_scores <- ifn_expression %>%
    filter(gene_name %in% ifn_davenport) %>%
    group_by(sampleid) %>%
    summarise(ifn_daven = sum(stdscaled)) %>%
    ungroup()


ifn_df <- left_join(ifn_a_scores, ifn_g_scores) %>%
    left_join(ifn_davenp_scores)

```



## PCA

PCA was computed for genes with TPM > 1 in at least 50% of individuals (about 10,000 genes).


```{r}
pca <- prcomp(expressed_matrix, center = TRUE, scale. = TRUE, rank. = 100)

pc_scores <- as_tibble(pca$x, rownames = "sampleid")
```

```{r, fig.height = 4, fig.width = 8}
pc1.2 <- ggplot(pc_scores, aes(PC1, PC2)) +
    geom_point(size = 2.5) +
    theme_minimal()

pc3.4 <- ggplot(pc_scores, aes(PC3, PC4)) +
    geom_point(size = 2.5) +
    theme_minimal()

plot_grid(pc1.2, pc3.4, nrow = 1)
```

### Variance explained

```{r, fig.height=3, fig.width=6}
tibble(PC = 1:length(pca$sdev),
       v = pca$sdev^2) %>%
mutate(pct = v/sum(v)) %>%
filter(PC %in% 1:50) %>%
    ggplot(aes(PC, pct)) +
    geom_point() +
    geom_line(aes(group = 1)) +
    scale_y_continuous(labels = scales::percent,
		       breaks = scales::pretty_breaks(6)) +
    theme_minimal() +
    labs(y = "% variance")
```



### Gene weights

```{r, fig.height = 6}
pc_loadings <- as_tibble(pca$rotation, rownames = "gene_id") %>%
    pivot_longer(-gene_id, names_to = "pc") %>%
    mutate(s = ifelse(value > 0, 1, 0)) %>%
    group_by(pc, s) %>%
    slice_max(n = 25, order_by = abs(value)) %>%
    ungroup() %>%
    left_join(gene_ids, by = "gene_id") %>%
    select(gene_id, gene_name, pc, value)

pc_loadings %>%
    filter(pc %in% c("PC1", "PC2")) %>%
    ggplot(aes(x = value, y = reorder_within(gene_name, by = value, within = pc))) +
    geom_col(aes(fill = value > 0), show.legend = FALSE) +
    scale_y_discrete(labels = function(x) str_remove(x, "_+PC\\d+$")) +
    facet_wrap(~pc, scales = "free") +
    theme_minimal() +
    labs(x = "PC loadings", y = NULL)
```

### PCA colored by IFN score

IFN score was computed as the cummulative scaled expression levels for IFN genes.

- "Interferon Alpha Genes" are genes included in GSEA's HALLMARK INTERFERON ALPHA list. 
- "Interferon Gamma Genes" are genes included in GSEA's HALLMARK INTERFERON GAMMA list. 
- "Interferon Davenport Set" are 11 genes listed in Davenport et al (2018).

```{r, fig.height = 3, fig.width = 10}
pc_scores %>%
    select(sampleid, PC1:PC2) %>%
    left_join(ifn_df, by = "sampleid") %>%
    pivot_longer(starts_with("ifn")) %>%
    mutate(name = recode(name, "ifn_a" = "Interferon Alpha Genes",
			 "ifn_g" = "Interferon Gamma Genes",
			 "ifn_daven" = "Interferon Davenport Set"),
	   name = fct_inorder(name)) %>%
    split(.$name) %>%
    map(~ggplot(., aes(PC1, PC2)) +
    geom_point(aes(fill = value), size = 2.5, shape = 21) +
    scale_fill_viridis_c(option = "magma",
			 guide = guide_colorbar(barwidth = .5)) +
    facet_wrap(~name) + 
    theme_minimal() +
    theme(panel.grid.minor = element_blank()) +
    labs(fill = "IFN\nscore")) %>%
    plot_grid(plotlist = ., nrow = 1)
```

### Correlation between IFN score and PC1 


#### Interferon Alpha Genes

```{r}
test_df <- pc_scores %>%
    select(sampleid, PC1) %>%
    left_join(ifn_df, by = "sampleid")

cor.test(test_df$ifn_a, test_df$PC1, method = "spearman")
```

#### Interferon Gamma Genes

```{r}
cor.test(test_df$ifn_g, test_df$PC1, method = "spearman")
```

#### Interferon Davenport Genes

```{r}
cor.test(test_df$ifn_daven, test_df$PC1, method = "spearman")
```



### K-means clustering of IFN genes expression levels 

```{r, fig.height = 3, fig.width = 10}
ifn_a_clusters <- ifn_expression %>%
    filter(gene_name %in% ifn_a) %>%
    select(sampleid, gene_name, stdexp) %>%
    pivot_wider(names_from = gene_name, values_from = stdexp) %>%
    column_to_rownames("sampleid") %>%
    as.matrix() %>%
    kmeans(2) %>%
    .$cluster %>%
    as.data.frame() %>%
    rownames_to_column("sampleid") %>%
    select(sampleid, cluster = 2) %>%
    as_tibble()

ifn_g_clusters <- ifn_expression %>%
    filter(gene_name %in% ifn_g) %>%
    select(sampleid, gene_name, stdexp) %>%
    pivot_wider(names_from = gene_name, values_from = stdexp) %>%
    column_to_rownames("sampleid") %>%
    as.matrix() %>%
    kmeans(2) %>%
    .$cluster %>%
    as.data.frame() %>%
    rownames_to_column("sampleid") %>%
    select(sampleid, cluster = 2) %>%
    as_tibble()

ifn_daven_clusters <- ifn_expression %>%
    filter(gene_name %in% ifn_davenport) %>%
    select(sampleid, gene_name, stdexp) %>%
    pivot_wider(names_from = gene_name, values_from = stdexp) %>%
    column_to_rownames("sampleid") %>%
    as.matrix() %>%
    kmeans(2) %>%
    .$cluster %>%
    as.data.frame() %>%
    rownames_to_column("sampleid") %>%
    select(sampleid, cluster = 2) %>%
    as_tibble()

ifn_cluster_df <- 
    bind_rows("Interferon Alpha Genes" = ifn_a_clusters,
	      "Interferon Gamma Genes" = ifn_g_clusters,
	      "Interferon Davenport Set" = ifn_daven_clusters,
	      .id = "set") %>%
    mutate(set = fct_inorder(set))

pc_scores %>%
    select(sampleid, PC1:PC2) %>%
    left_join(ifn_cluster_df, by = "sampleid") %>%
    ggplot(aes(PC1, PC2)) +
    geom_point(aes(fill = factor(cluster)), size = 2.5, shape = 21) +
    scale_fill_viridis_d() +
    facet_wrap(~set, nrow = 1) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank()) +
    labs(fill = "IFN\ncluster")
```

```{r, eval=FALSE}
write_tsv(ifn_g_clusters, "./clusters_IFNg.tsv")
```






